---
  html_document:
    toc: true
    highlight: zenburn
    theme: united
---
The code for this report is on github [here](https://github.com/hbc/rubin-cell-fate-conversion).

```{r setup, echo=FALSE}
knitr::opts_chunk$set(tidy=TRUE, highlight=TRUE, dev="png",
               cache=TRUE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='')
```
```{r load-libaries}
library(useful)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
```

Read in the data for each plate into one big dataframe, make unique IDs for
each batch/well combination and use those IDs as the column names.

```{r read-plates}
plate_reader = function(fn) {
  batch = strsplit(basename(fn), ".", fixed=TRUE)[[1]][1]
  data = read.table(fn, header=TRUE, row.names=1)
  colnames(data) = paste(batch, colnames(data), sep="_")
  data
}
plates_fns = sort(list.files("data", pattern="*.dat", full.names=TRUE))
plates = do.call(cbind, lapply(plates_fns, plate_reader))
```

Create a dataframe of the metadata about each sample. This has an identifier
for a sample, Which well it came from, which batch and what it was treated with.

```{r massage-well-identity-data}
welldata_fn = "metadata/Compound Layout 384w.csv"
welldata = read_csv(welldata_fn) %>%
  gather(column, treatment, -row) %>%
  mutate(well=paste(row, column, sep="")) %>%
  dplyr::select(well, treatment)
identities = data.frame(str_split_fixed(colnames(plates), "_", 3))
colnames(identities) = c("batch", "drop", "well")
identities$id = colnames(plates)
welldata = identities %>% dplyr::select(batch, well) %>%
         left_join(welldata, by="well")
rownames(welldata) = colnames(plates)
welldata$sample = colnames(plates)
welldata$classes = ifelse(welldata$treatment == "DMSO", "DMSO",
      ifelse(welldata$treatment == "AdMyoD", "AdMyoD", "other"))
```


Verify that the samples match up between the read counts and the metadata dataframe
and make sure there are no NA counts.

```{r verify-data}
dim(welldata)
dim(plates)
table(rownames(welldata) %in% colnames(plates))
corner(plates)
corner(welldata)
table(complete.cases(plates))
```

Looks like we are good to go. While we're at it we will load the positive
control data.

```{r load-positive-control-data}
positive_fn = "data/Feo_positive_controls.unq.refseq.umi.SC_and_Myoblast_Raw.txt"
positive = read.table(positive_fn, header=TRUE, row.names=rownames(plates))
positive$id = NULL
positive_samples = data.frame(str_split_fixed(colnames(positive), "_", 3))
identities = data.frame(str_split_fixed(colnames(plates), "_", 3))$X1
positive_welldata = data.frame(batch=positive_samples$X1,
                               treatment=positive_samples$X1,
                               well=positive_samples$X1)
```

Now we will calculate some summary statistics about each sample.

```{r plot-genes-detected}
welldata$genes_detected = colSums(plates > 0)
welldata$genes_detected_zscore = ave(welldata$genes_detected, FUN=scale)
welldata$genes_detected_pval = 2 * pnorm(-abs(welldata$genes_detected_zscore))
welldata$genes_detected_padj = p.adjust(welldata$genes_detected_pval, method="BH")
```

We see around 10k genes detected in each cell on each plate, with some cells with
a low number of genes detected on each plate.

```{r low-counts-plate}
ggplot(welldata, aes(batch, genes_detected)) + geom_boxplot() +
   ylab("genes with counts > 0") +
   xlab("") +
   theme_bw()
```

```{r boxplot-counts-plate}
welldata$counts = colSums(plates)
ggplot(welldata, aes(batch, counts)) + geom_boxplot() +
   ylab("total counts") +
   xlab("") +
   theme_bw()
```

```{r plot-complexity}
ggplot(welldata, aes(counts, genes_detected, color=batch)) +
   geom_smooth(fill=NA) +
   ylab("genes with counts > 0") +
   xlab("total counts") +
   theme_bw()
```

In this histogram of the genes detected, we can see there are a set of
cells with a low amount of genes detected, more in the M2 plate than
the other plates but for the most part the plates look pretty similar.

```{r genes-detected-histogram}
ggplot(welldata, aes(genes_detected)) + geom_histogram() +
  theme_bw() + xlab("genes with counts > 0") + facet_wrap(~batch)
```

Here we drop the cells with less than 7,500 genes detected.

```{r drop-low-genes-detected}
welldata = subset(welldata, genes_detected > 7500)
plates = plates[, rownames(welldata)]
```

```{r biomart-conversions}
library(biomaRt)
mouse = useMart(biomart = "ENSEMBL_MART_ENSEMBL",
                dataset="mmusculus_gene_ensembl",
                host = "jul2015.archive.ensembl.org")
conversions = getBM(attributes=c("ensembl_gene_id", "mgi_symbol", "gene_biotype"),
                    mart=mouse)
```

Samples do not have very many counts in noise genes, so that is not an issue.
Noise genes are flagged by either being too small to be picked up reliably in a
standard RNA-seq analysis or are highly variable and prone to introducing
distortion such as rRNA.

```{r add-noise-counts}
biotypes = unique(conversions$gene_biotype)
noise_rna_biotypes = c("Mt_tRNA", "Mt_rRNA", "snoRNA", "snRNA", "misc_RNA",
                       "scaRNA", "rRNA", "sRNA")
noise_rna_genes = subset(conversions, gene_biotype %in% noise_rna_biotypes)$mgi_symbol
noise_rna = rownames(plates)[rownames(plates) %in% noise_rna_genes]
welldata$noise_counts = colSums(plates[noise_rna,])
ggplot(welldata, aes(batch, noise_counts)) + geom_boxplot() +
   ylab("counts in noise genes") +
   xlab("") +
   theme_bw()
```

We'll drop the noise genes from consideration even though there aren't many
counts in them.

```{r drop-noise-genes}
plates = plates[!rownames(plates) %in% noise_rna_genes,]
```

We'll also drop all genes with that don't have at least 100 counts total and
are not seen in at least 4 samples. This cuts down the number of genes we
are considering to ~13,500.

```{r drop-low-genes}
plates = plates[rowSums(plates > 0) > 4 & rowSums(plates) > 100,]
```

This is what we're left with in terms of samples:

```{r filtering-finish}
knitr::kable(welldata %>% group_by(batch) %>% summarize(total=n()))
```

and we're left with `r nrow(plates)` genes to consider.

```{r seurat-setup}
library(Seurat)
seurat.data = new('seurat', raw.data=plates)
seurat.data = Setup(seurat.data, project="rubin", min.cells=3, min.genes=1000,
                    meta.data=welldata, total.expr=1e4)

```

Here we look at what are the most variable genes across the samples. We can
see a lot of these are subunits of ribosomal proteins, these will be used
for the PCA.

```{r mean-var-plot}
seurat.data = MeanVarPlot(seurat.data, y.cutoff = 0.5, x.low.cutoff = 0.0125,
                          x.high.cutoff=3,
                          fxn.x = expMean,fxn.y = logVarDivMean)
```

We'll focus on component 1 since that seems to be what separates out
the batch 3 plate from the other plates.

```{r pca-plot}
seurat.data = PCA(seurat.data, do.print=FALSE)
PCAPlot(seurat.data, 1, 2, pt.size=2)
rot = seurat.data@pca.rot %>%
  tibble::rownames_to_column(var="sample") %>%
  left_join(welldata, by="sample")
ggplot(rot, aes(PC1, PC2, shape=classes, color=batch)) +
  geom_point(size=2) +
  theme_bw()
```

There are quite a few ribosomal proteins that are flagged. These should for the
most part eithr be stably expressed or are not particularly interesting hits.

Since TCL3 is so different, and it seems different in non-interesting
ways, let's just drop it for now to make our lives easier. We can see
that after dropping TCL3 we see a separation of the AdMyoD samples from
the other samples.

```{r drop-tcl3}
welldata = subset(welldata, batch != "M3")
plates = plates[, rownames(welldata)]
seurat.data = new('seurat', raw.data=plates)
seurat.data = Setup(seurat.data, project="rubin", min.cells=3, min.genes=200,
                    total.expr=1e4, meta.data=welldata)
seurat.data = MeanVarPlot(seurat.data, y.cutoff = 0.5, x.low.cutoff = 0.0125,
                          x.high.cutoff = 3, do.contour=F,
                          fxn.x = expMean,fxn.y = logVarDivMean)
seurat.data = PCA(seurat.data, do.print=FALSE)
PCAPlot(seurat.data, 1, 2, pt.size=2)
seurat.data = ProjectPCA(seurat.data, do.print = FALSE)
VizPCA(seurat.data, pcs.use = 1:3, num.genes = 10, use.full = TRUE, nCol = 3)
rot = seurat.data@pca.rot %>%
  tibble::rownames_to_column(var="sample") %>%
  left_join(welldata, by="sample")
ggplot(rot, aes(PC1, PC2, color=batch, label=treatment)) +
  geom_text(size=2) +
  theme_bw()
```

We see there are some chemicals that seem to move the cells towards the AdMyoD
treated cells. They seem reproducible across the different batches, which is
great. Zooming in on that area shows some more candidates:

```{r pca-zoom}
ggplot(rot, aes(PC1, PC2, color=batch, label=treatment)) +
  coord_cartesian(xlim=c(-30, 0), ylim=c(-25, 0)) +
  geom_text(size=2) +
  theme_bw()
```

# Distance to AdMyoD samples
Here we take a different tactic and find the genes that are most different
between the AdMyoD samples and the DMSO treated samples. Then we will use those
genes to measure how far each of the treated samples is from the AdMyoD samples
in terms of expression of those genes.

```{r admyo-distance-calculation}
md_samples = rownames(subset(welldata, treatment %in% c("AdMyoD", "DMSO")))
md_metadata = welldata[md_samples,]
md_counts = plates[, md_samples]
```

We fit a model that tests for differences between AdMyoD and DMSO treatment
while controlling for batch effects.

```{r deseq2-model-fit}
library(DESeq2)
dds = DESeqDataSetFromMatrix(countData=md_counts, colData=md_metadata,
                             design=~batch+treatment)
dds = DESeq(dds)
plotDispEsts(dds)
res = results(dds)
sig = subset(res, padj < 0.05) %>%
  data.frame() %>%
  tibble::rownames_to_column(var="gene") %>%
  dplyr::arrange(padj)
```

There are `r nrow(subset(res, padj < 0.05))` genes tagged as differentially
expressed between the AdMyoD and the DMSO treated samples. We'll use those
genes to measure how far the chemical treated samples are from the AdMyoD treated
samples. This gives different results than the PCA method we were using:

```{r euclidean-distance}
comp1class = welldata[, c("sample", "treatment", "classes")]
colnames(comp1class) = c("sample", "comp1treat", "comp1class")
comp2class = welldata[, c("sample", "treatment", "classes")]
colnames(comp2class) = c("sample", "comp2treat", "comp2class")
dds = DESeqDataSetFromMatrix(countData=plates, colData=welldata,
                             design=~batch+treatment)
dds = estimateSizeFactors(dds)
ncounts = log(counts(dds, normalized=TRUE) + 1)
dists = as.matrix(dist(t(ncounts[sig$gene,])))
dists[diag(dists)] = NA
dists = dists %>%
  data.frame() %>%
  na.omit() %>%
  tibble::rownames_to_column() %>%
  tidyr::gather(sample, distance, -rowname)
colnames(dists) = c("comp1", "comp2", "distance")
dists = dists %>%
  left_join(comp1class, by=c("comp1"="sample")) %>%
  left_join(comp2class, by=c("comp2"="sample")) %>%
  group_by(comp1treat, comp2treat) %>%
  summarise(mtreatdist=mean(distance), mtreatsd=sd(distance))
ggplot(subset(dists, comp1treat == "AdMyoD"),
       aes(comp2treat, mtreatdist, label=comp2treat)) +
  geom_text() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
write_csv(subset(dists, comp1treat == "AdMyoD"), "normalized-count-distance-to-admyod.csv")
```

## PCA distances
We can also calculate the distances between points on PC1 and PC2 to get an
estimate of how similar the samples are to each other:

```{r pca-distances)
pcamat = as.matrix(rot[, c("PC1", "PC2")])
rownames(pcamat) = rot$sample
pcadist = as.matrix(dist(pcamat))
pcadist[diag(pcadist)] = NA
pcadist = pcadist %>%
  data.frame() %>%
  na.omit() %>%
  tibble::rownames_to_column() %>%
  tidyr::gather(sample, distance, -rowname)
colnames(pcadist) = c("comp1", "comp2", "distance")
pcadist = pcadist %>%
  left_join(comp1class, by=c("comp1"="sample")) %>%
  left_join(comp2class, by=c("comp2"="sample")) %>%
  group_by(comp1treat, comp2treat) %>%
  summarise(mtreatdist=mean(distance), mtreatsd=sd(distance))
myod = subset(pcadist, comp1treat == "AdMyoD") %>%
  dplyr::arrange(mtreatdist) %>% mutate(rank=dense_rank(mtreatdist))
ggplot(myod,
       aes(rank, mtreatdist, label=comp2treat)) +
  geom_point() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  ylab("Euclidean distance to AdMyoD")
write_csv(myod, "pca-distance-to-admyod.csv")
```
